---
title: "<span style='font-size: 14px;'>BAMexploreR: model-based distribution, abundance, and habitat associations of boreal birds</span><br><strong>Model Access</strong>"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BAMexploreR_2_access}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "",
  message = FALSE,
  warning = FALSE
)
```


This vignette shows you how to access the BAM Landbird Density & Habitat models using the `BAMexploreR` package.

Throughout this vignette, `version = "v5"` selects the current models (Version 5), while `version = "v4"` selects the archived models (Version 4).

## Background

### Model subregions
The BAM Landbird Density & Habitat models are produced by modelling multiple ecologically distinct regions and mosaicking the model predictions into a seamless layer.

- **Archived-model regions (Version 4)**: Bird Conservation Regions (BCRs) in Canada.
- **Current-model regions (Version 5)**: The intersection of national boundaries and updated Bird Conservation Regions (BCRs).

You can choose to access predictions for the entire modelled area or a custom set of model subregions using an area of interest. The archived models have one available mosaic for Canada, while the current models have three available mosaics: Canada, Alaska, and the Lower 48 United States. These regions are provided as separate mosaics because the models use the best available country-specific remote-sensing products as predictors. For more information on the BAM Landbird Density & Habitat models, see the 'Introduction' vignette.

### Subregion naming convention
Because the current models use the intersections of countries and BCRs as modelling subregions, each subregion is named using a country prefix and BCR number. BCR labels differ between the archived and current models because several BCR boundaries were updated. Several BCRs were also divided further in Version 5, as indicated by the hyphenated numbering system.

```{r list_names, message=FALSE, warning=FALSE, echo=FALSE}
library(BAMexploreR)
cat("Version 4 subregion names: ", bam_get_bcr("v4"), "\n")
cat("  \n")
cat("Version 5 subregion names: ", bam_get_bcr("v5"), "\n")

```
## Functions available

### bam_map_bcr()

**bam_map_bcr()** plots the model subregions for a chosen version and can optionally overlay a provided SpatVector to highlight the area of interest.

Arguments

- version: Specifies the model release: `"v5"` for the current models or `"v4"` for the archived models.

- ext: SpatVector or SpatRaster used to define an area of interest.


Output

Map illustrating the BCR and overlap extent if provided.


### bam_get_bcr()

**bam_get_bcr()** returns the list of model subregions for a given model version. If an ext object is supplied, only those subunits that intersect the provided extent are returned.

Arguments

- version: Specifies the model release: `"v5"` for the current models or `"v4"` for the archived models.

- ext: (Optional) SpatVector or SpatRaster whose extent will be used to filter and return only intersecting BCR subunits.


Output

A vector representing the name of the BCR subunits. 


### bam_get_layer()

The **bam_get_layer()** function downloads BAM Landbird Density & Habitat model rasters for specified species and areas of interest.

Arguments

- spList: A named list of SpatRaster objects, each representing predicted density for a species.

- version: Specifies the model release: `"v5"` for the current models or `"v4"` for the archived models.

- destfile: A character indicating output path where the downloaded file is saved.

- crop_ext: A SpatVector or SpatRaster used to define the extent for the cropping.

- year: A character specifying the prediction year for the current models (`version = "v5"`). Only `"2020"` is currently available for public download and is the default. Predictions at five-year intervals from 1995 to 2015 are available by request from [bamp@ualberta.ca](mailto:bamp@ualberta.ca).

- bcrNM: A vector representing the BCR subunit name according to model version. Default is "mosaic".


Output

A list of multiband SpatRaster objects. In addition to returning these objects, the function also downloads raster files to the directory specified by destfile as a side-effect.

## Example workflow

### 1. Identify species available

The first step in accessing the BAM Landbird Density & Habitat model products is determining the species of interest.

There are 143 species available for the archived models and 149 species available for the current models. You can retrieve the species available for a specified model release using `bam_spp_list()`. The function's `guild` parameter can filter the results using the classifications from The State of Canada's Birds, which categorizes species by broad biome and by groups with distinct and noteworthy trends. Species assessed as Special Concern, Threatened, or Endangered by the Committee on the Status of Endangered Wildlife in Canada (COSEWIC) are also identified. If `guild` is not provided, the function returns all species available for the specified model release.

Let's check the species group categories by accessing the `guild_opt` object.

```{r species_groups, echo = TRUE, results = 'markup', error = TRUE}
# Get the options for species grouping
guild_opt
```
In our example, we'll retrieve four-letter AOU codes for COSEWIC-listed species available in the current models (`version = "v5"`).

```{r species_list_code, echo = TRUE, results = 'markup', error = TRUE}
# Generate a vector of COSEWIC species codes available in version 5. 
cosewic_v5 <- bam_spp_list("v5", "speciesCode", "COSEWIC")
cosewic_v5
```

We can also retrieve the common names for our COSEWIC species.

```{r species_list_common, echo = TRUE, results = 'markup', error = TRUE}
# Generate a vector of COSEWIC species common names available in version 5. 
cosewic_v5_cn <- bam_spp_list("v5", "commonName", "COSEWIC")
cosewic_v5_cn
```

### 2. Identify model subregions that intersect the area of interest (AOI)

Next, we'll need to determine the AOI for accessing the model products. 

You can access predictions for the entire modelled area, a single model subregion (current models only), or a custom AOI. Because model subregion names vary by model release and country, `bam_get_bcr()` and `bam_map_bcr()` ensure that the correct names are used. You can specify an extent to identify the BCRs that intersect the AOI.

For our example, we'll use an example AOI that intersects two model subregions in the boreal plains and shield. Let's get the list of model subregions that interest with our AOI.

```{r aoi_bcr_list, echo = TRUE, results = 'markup', error = TRUE}
library(terra)

# Use the example aoi SpatVector provided in the package
aoi_sf <- vect(system.file("extdata", "vignette_poly_3978.shp", package = "BAMexploreR"))

# Extract subregions from version 5 that intersect the aoi
bcraoi_v5 <- bam_get_bcr("v5", ext=aoi_sf) 
bcraoi_v5
```

We can also use `bam_map_bcr()` to visualize where the AOI falls within the current model subregions.

```{r aoi_bcr_map, echo = TRUE, results = 'markup', error = TRUE, fig.width=8, fig.height=4}
# Map aoi on the map of model subregions
map <- bam_map_bcr("v5", ext=aoi_sf) +
  tmap::tm_title("Model Subregions - Version 5", position = c("center", "top"))
suppressWarnings(print(map))
```

### 3. Download species specific density maps

Now that we have our species and AOI, we can download the model products!

We will download the model products using the `bam_get_layer` function. 

Let's download a current-model `SpatRaster` for one of our COSEWIC-designated species and the two BCRs that intersect the AOI.


```{r get_layer, echo = TRUE, message=FALSE, results='hide'}
#Download the model product for Olive-sided Flycatcher
result <- bam_get_layer(cosewic_v5[7], "v5", destfile = tempdir(), year = "2020", bcrNM=bcraoi_v5)
```

The output is a list of `SpatRaster` objects. This example returns a list of one object because we requested one species. Each current-model object has two bands:

- **mean**: the mean predicted density in males/ha across 32 bootstraps
- **cv**: the coefficient of variation of predicted density across 32 bootstraps

The archived model products contain only the **mean** band.

Finally, we can plot these bands for our species of interest, Olive-sided Flycatcher.

```{r plot_layer, echo = TRUE, results = 'markup', error = TRUE, fig.width=8, fig.height=3}
library(terra)

#basic plotting of all bands
plot(result$OSFL)
```
