---
title: "3. BAMexploreR Model Covariates Analysis"
author: "Diego, Siu Chung WU"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{model_covariates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "",
  message = FALSE,
  warning = FALSE
)
```

```{r setup_packages, echo = FALSE, results = 'hide'}
library(BAMexploreR)
suppressWarnings(suppressMessages(library(httr)))
library(tmap)
library(dplyr)
library(stringr)
library(terra)
library(tools)
tmap::tmap_mode("plot")
data("bam_covariate_importance_v4", package = "BAMexploreR")
```

# Introduction

`BAMexploreR` is a package for analyzing bird species data from the BAM National Model. This vignette will guide you through the usage of covariate function in the package, especially for comparing importance of covariates, and their interactions.


# Covariates analysis: stacked bar plot

Below is the sample function to create a stacked bar plot (y = variable class, x= importance) on the list of covariates used in the model, for the model version 4.
```{r stacked_bar, echo = TRUE, results = 'markup', error = TRUE}
stackedbarNM(data = bam_covariate_importance_v4, species = "all", bcr = "all", group = c("var_class", "bcr"), plot = FALSE)
```

# Covariates analysis: box plot

Below is the sample function to create a box plot




