---
title: "<span style='font-size: 14px;'>BAM National Models Density Map</span><br><strong>Accessing the data</strong>"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{model_distribution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "",
  message = FALSE,
  warning = FALSE
)
```



## Overview
The BAMexploreR package contains functions to explore bird :
vignette demonstrates how to use the **occurrenceNM()** and **pop_sizeNM()** functions to estimate species presence and population size from spatial raster data.

These functions work with downloaded a list of raster acquired using getLayer function.


### occurrenceNM(): Estimate Presence Using Abundance Thresholds

Description
The occurrenceNM() function takes one or more abundance rasters (one per species) and determines a presence-absence threshold for each using the Lorenz curve method. The function then estimates the proportion of area where the species is present.

Arguments
- raster_list: A list of SpatRaster objects (one per species).

- quantile: Method to derive the threshold; default is "by_lorenz".

- plot: Logical, if TRUE returns a ggplot object showing the Lorenz curve and selected threshold.

📤 Output
A tibble with the following columns:

- species: Name or index of the species.

- threshold: Abundance value used to classify presence.

- presence_prop: Proportion of area where abundance exceeds the threshold.


## Installation

You can install the development version of `BAMexploreR` from [GitHub](https://github.com/) with:

```{r setup, include = TRUE, message=FALSE, warning=FALSE}
library(BAMexploreR)
```

### Data
All data used in this example are included in the package and are described in the help files.
```{r, eval = FALSE}
?spp_List

```



## Example workflow

### Mapping species occurence 

```{r get_layer, fig.width=8, fig.height=6, echo = TRUE, results = 'markup', error = TRUE}
library(terra)

aoi_sf <- vect(system.file("extdata", "vignette_poly_5072.shp", package = "BAMexploreR"))

# Download TEWA and CAWA distribution maps and calculate population size within the area of interest 
result <- getlayerNM("TEWA", "v5", destfile = tempdir(), crop_ext = aoi_sf, year = "2010")
occ_results <- occurrenceNM(result)
```

## Citation

Please cite the _BAMexploreR_ package when using it in publications. To cite the latest official version, please use:


```{r, echo = FALSE, results = "asis"}
dev_version <- packageDescription("BAMexploreR")$Version
cat(paste0("\\> Houle M, Boehm M, Wu SC, Knight E, ... (", format(Sys.time(), "%Y"),"). BAMexploreR: Evaluating boreal birds density in Canada using R. R package version ", dev_version, ". Available at https://github.com/borealbirds/BAMexploreR\n"))
```



