---
title: "<span style='font-size: 14px;'>BAM Landbird Density Models</span><br><strong>Mapping Species Occurrence and Density</strong>"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{model_distribution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "",
  message = FALSE,
  warning = FALSE
)
```


The BAMexploreR package provides tools to explore boreal bird species distribution and estimate population metrics using raster outputs from predictive models. The following core functions are designed to support conservation planning and ecological analysis:


### bam_occurrence()

The **bam_occurrence()** function estimates a species' presence/absence threshold based on a Lorenz curve, which summarizes the spatial inequality in predicted occurrence probabilities. This function helps distinguish between areas of high and low predicted use, by determining a threshold value that defines the most likely occupied areas.

Arguments

- raster_list: A named list of SpatRaster objects, each representing predicted density for a species.

- quantile: Method to derive the threshold; default is "by_lorenz".

- plot: Logical, if TRUE returns a map with binary presence (1) / absence (0) values.

📤 Output

A tibble with the following columns:

- species: Species code identifying the species.

- type: Indicates whether a presence/absence threshold was applied when calculating the area.Values are "no_threshold" or "with_threshold".

- area_km2: The total area (in square kilometers) where predicted abundance exceeds the threshold (if applied).


### bam_pop_size()

The **bam_pop_size()** function estimates the total population size by summarizing predicted densities of males across a spatial extent. This function supports cropping to a defined area (e.g., a region of interest) and group-based summaries if the extent includes a grouping variable (e.g., subregions, jurisdictions).

Arguments

- raster_list: A named list of SpatRaster objects, each representing predicted density for a species.

- crop_ext: A SpatVector defining the spatial extent for the cropping and grouping of population estimates.

- group: (Optional) Name of the column in crop_ext to use for grouping population estimates.

📤 Output

A tibble with the following columns:

- group: Group identifier, such as a region or subunit (set to NA if no grouping was applied).

- total_pop: Estimated total number of males across the area (inferred from pixel-level densities).

- mean_density: Average predicted density (males per pixel).

- sd_density: Standard deviation of predicted densities.

- n_cells: Number of raster cells included in the calculation.

- species: Species code identifying the species.



## Example workflow

### Set  the area of interest (AOI) as the Canadian extent of BCRs 61 and 62 

```{r select_aoi, fig.width=8, fig.height=6, echo = TRUE, results = 'markup', error = TRUE}
library(BAMexploreR)
library(terra)
library(dplyr)

# Load and extract Canada BCRs 60 and 61
bcr_v5 <- vect(system.file("extdata", "BAM_BCRNMv5_5072.shp", package = "BAMexploreR"))
aoi_sf <- bcr_v5[bcr_v5$subunit_ui %in% c("can60", "can61"), ]

plot(bcr_v5, main = "AOI defined by the Canadian extent of BCRs 60 and 61")
plot(aoi_sf, col = "red", add = TRUE)
 

```

### Mapping species occurence for CAWA and OVEN within the AOI


```{r map_occurence, fig.width=8, fig.height=6, message = FALSE, warning= FALSE, echo = TRUE, results = 'markup', error = TRUE}
# Download TEWA and BBWO distribution maps and map presence/absence through the aoi 
buildresults <- bam_get_layer(c("CAWA", "OVEN"), "v5", destfile = tempdir(), crop_ext = aoi_sf, year = "2010")
occ_results <- bam_occurrence(results, plot = TRUE)

```


### Estimating species specific population size within the AOI per BCR


```{r map_density, fig.width=8, fig.height=6, message = FALSE, warning= FALSE, echo = TRUE, error = TRUE}
library(knitr)

# Calculate population size based on BAM National Model density raster
occ_results <- bam_pop_size(results, crop_ext = aoi_sf, group = "subunit_ui")

# Show results
kable(occ_results, caption = "Estimated population size grouped by BCR")
```

## Citation

Please cite the _BAMexploreR_ package when using it in publications. To cite the latest official version, please use:


```{r, echo = FALSE, results = "asis"}
dev_version <- packageDescription("BAMexploreR")$Version
cat(paste0("\\> Houle M, Boehm M, Wu SC, Knight E, ... (", format(Sys.time(), "%Y"),"). BAMexploreR: Evaluating boreal birds density in Canada using R. R package version ", dev_version, ". Available at https://github.com/borealbirds/BAMexploreR\n"))
```



