---
title: "Using BAMexploreR"
subtitle: "Accessing data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using BAMexploreR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(BAMexploreR)
devtools::load_all()
library(tmap)
```

## Introduction

`BAMexploreR` is a package for downloading and analyzing bird species data from the BAM National Model v4 and v5. Functions in the package perform the following tasks:
* Accessing the data - list the species for which National models were created, specify and map bcr unit available for each National model version and download species output.
* Population & distribution - 
* Covariate interpretation - 
 
This vignette will guide you through the basic usage of accessing the data using the package, including how to download specific species using region subunit or custom extent.

## Example workflow
The following workflow demonstrates 

### Data
All data used in this example are included in the package and are described in the help files.
```{r, eval = FALSE}
?guild_opt
?sppList
?vignette_intact
?vignette_led
?vignette_streams
```



## BCR naming convention
Because most of the spatial layers used to develop the models were country specific, we used as model-building units the combination of bird conservation regions (BCRs) intersected with Canada-USA jurisdictional boundaries. Those subunits changed overtime based on decision making by the different agencies. Each subunits use acronym of the country as prefix follwed by the nmber of the BCR unit.  








